---
title: "Defining sets of management actions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Defining sets of management actions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(multiscape)
```

An action set gives a name to a combination of existing management actions.
For example, restoration and invasive-species control can be members of
`restore_control`, while the same restoration action can also belong to
`restore_fence`. The actions remain individual decisions; membership is
many-to-many.

## Register the individual actions first

```{r define-actions}
problem <- create_problem(
  pu = data.frame(id = 1:2, cost = c(1, 2)),
  features = data.frame(id = 1L, name = "woodland"),
  dist_features = data.frame(
    pu = 1:2, feature = 1L, amount = c(10, 20)
  )
) |>
  add_actions(
    actions = data.frame(id = c("restore", "control", "fence")),
    cost = c(restore = 2, control = 1, fence = 1)
  )

# No combinations have been registered yet.
get_action_sets(problem)
```

## Use a named list

The list names are set identifiers; each vector contains registered action ids.

```{r list-input}
list_problem <- add_action_sets(problem, list(
  restore_control = c("restore", "control"),
  restore_fence = c("restore", "fence")
))
get_action_sets(list_problem)
```

## Use a long table

Each row defines one membership. An action can occur in several rows as long as
it belongs to different sets.

```{r table-input}
memberships <- data.frame(
  set = c("restore_control", "restore_control", "restore_fence", "restore_fence"),
  action = c("restore", "control", "restore", "fence")
)
table_problem <- add_action_sets(problem, memberships)
get_action_sets(table_problem)
identical(get_action_sets(list_problem), get_action_sets(table_problem))
```

Both formats are stored as a long table and returned in a consistent order by
`get_action_sets()`. The input problem is preserved:

```{r original-problem}
get_action_sets(problem)
```

## Add further combinations

Subsequent calls add new sets without replacing existing definitions:

```{r additional-set}
extended_problem <- table_problem |>
  add_action_sets(list(control_fence = c("control", "fence")))
get_action_sets(extended_problem)
get_action_sets(table_problem)
```

Each set must contain at least two distinct actions. A set identifier cannot
match an action id, an action display name, or an existing `actions$action_set`
classification label. Members must be action ids; sets cannot be nested.
Duplicated memberships and reusing an existing set identifier are errors.
Updating the action catalog also checks that registered definitions remain
valid, so removing a member cannot silently leave a dangling set.

## What registration means

Registering a set does not create another action, modify costs or feasibility,
impose dependencies or exclusivity, or introduce a joint effect. The original
`actions$action_set` classification is also preserved. The same definitions can
therefore describe candidate combinations without deciding their management
rules.

```{r unchanged-actions}
identical(problem$data$actions, table_problem$data$actions)
identical(problem$data$dist_actions, table_problem$data$dist_actions)
```

In this release these are definitions only. The default is still at most one
selected action per planning unit. The separate
`add_constraint_action_cardinality()` function can allow concurrent actions
in cost/profit workflows; see the [action-cardinality vignette](Action_cardinality.html).
Set identifiers cannot yet be used as action inputs to effects, objectives, or
constraints. Joint ecological effects and feature aggregation remain separate
steps in the package's development.
