---
title: "Logical relations between management actions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Logical relations between management actions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(multiscape)
```

Cardinality controls how many actions can be selected. Logical relations
control which combinations are permitted **within each planning unit**.
They constrain existing individual decisions without adding selectable sets
or changing costs, profit, or effects.

| Function | Meaning in each scoped unit |
|:---------|:----------------------------|
| `add_constraint_action_requires(x, actions, requires, sense = "all", pu = NULL, name = NULL)` | Each selected trigger requires every companion. |
| `add_constraint_action_requires(..., sense = "any")` | Each selected trigger requires at least one companion. |
| `add_constraint_action_excludes(x, actions, pu = NULL, name = NULL)` | At most one group member can be selected. |
| `add_constraint_action_together(x, actions, pu = NULL, name = NULL)` | All group members are selected or none are. |

All three allow the scoped actions to remain unselected. They express neither
temporal order nor dependencies between different units. `name` is an optional
constraint label, independent of objective aliases used by MO methods.

## Define a common base

```{r base}
base <- create_problem(
  pu = data.frame(id = c(10L, 20L), cost = 5),
  features = data.frame(id = 1L, name = "woodland"),
  dist_features = data.frame(pu = c(10L, 20L), feature = 1L, amount = 100)
) |>
  add_actions(
    data.frame(id = c("restore", "control", "fence", "monitor")), cost = 1
  ) |>
  add_constraint_action_cardinality(3, "max")
```

The total maximum explicitly permits three simultaneous actions in both units.
Logical relations alone retain the implicit maximum of one. With that default,
a two-action together group cannot be selected, and a trigger requiring another
action cannot be selected either. Actions outside those groups remain possible.
Subset maxima and minima do not lift the default total maximum.

## Require all or any companions

```{r dependencies}
dependencies <- base |>
  add_constraint_action_requires(
    "restore", c("control", "monitor"), pu = 10L, name = "restore_requirements"
  ) |>
  add_constraint_action_requires(
    "restore", c("control", "fence"), sense = "any", pu = 20L,
    name = "alternative_protection"
  )
dependencies$data$constraints$action_relations
```

In unit 10, restoration needs **both** control and monitoring. In unit 20, it
needs **either** control or fencing (both also satisfy an any dependency).
Selecting a companion alone does not force restoration.

For binary decision $x_{i,a}$, all requires $x_{i,a} \le x_{i,b}$ for each
companion $b$. Any requires $x_{i,a} \le \sum_b x_{i,b}$. If `actions` contains
several triggers, **each trigger independently** obeys the relation; the model
does not wait for the entire triggering group to be selected. Triggering and
required groups must be disjoint.

## Exclude alternatives or implement a group together

```{r exclusions-and-together}
exclusive <- base |>
  add_constraint_action_excludes(c("control", "fence"), name = "protection_choice")

joint <- base |>
  add_constraint_action_together(c("restore", "control"), pu = 10L,
                                 name = "joint_implementation")
```

Excludes imposes $\sum_{a \in A}x_{i,a} \le 1$. For three or more members, every
pair is incompatible, rather than just forbidding the complete group.
Together equates all member decisions: selecting any member selects every
other member, while selecting none remains valid. Other actions can accompany
the group if cardinality and other constraints permit it.

## Use registered action-set members explicitly

```{r set-members}
sets <- base |>
  add_action_sets(data.frame(
    set = c("restore_control", "restore_control"),
    action = c("restore", "control")
  ))
membership <- get_action_sets(sets)
members <- membership$action[membership$set == "restore_control"]
sets_joint <- add_constraint_action_together(sets, members)
```

Registering a set alone does not impose a relation. These functions accept
individual action IDs (or legacy `actions$action_set` classification labels);
registered set IDs and display names are not resolved as decisions.

## Scope, unavailable actions, and repeated calls

`pu = NULL` resolves to all current external PU IDs. A vector such as
`pu = c(10L, 20L)` applies the rule separately in both units. A companion in
another unit cannot satisfy a dependency.

A missing feasible pair, locked-out action, or action removed for a non-finite
cost is treated as a zero decision. In an all dependency, one unavailable
companion prevents its trigger. In an any dependency, the remaining companions
can satisfy it; if none are available, the trigger is prevented. In a together
group, an unavailable member prevents every other member. Excludes simply
restricts the remaining members.

Distinct relations accumulate. Their identity includes type, action groups,
all/any sense, and PU scope; order and repeated input IDs do not matter.
With one companion, all and any have identical identity. Duplicate identities
raise an error even with a different name, and names must be unique across
the three relation functions. To compare different specifications, derive each
from the common base before adding that relation.

Cycles are allowed. For example, A requires B and B requires A imply joint
selection. Combining this with an A/B exclusion prohibits both; locking A in
then makes the model infeasible. The solver determines joint feasibility with
locks, budgets, cardinality, and all other constraints.

## Solve an economic example

This example uses independent action profits. For concurrent ecological effects,
see the [joint-effects vignette](Joint_effects.html). Restoring requires control;
their combined profit is 7, compared with fencing alone at 6.
The two-action capacity prevents adding fencing to the restoration group.

```{r economic-problem}
economic <- create_problem(
  pu = data.frame(id = 10L, cost = 5),
  features = data.frame(id = 1L, name = "woodland"),
  dist_features = data.frame(pu = 10L, feature = 1L, amount = 100)
) |>
  add_actions(data.frame(id = c("restore", "control", "fence")), cost = 1) |>
  add_constraint_action_cardinality(2, "max") |>
  add_constraint_action_requires("restore", "control") |>
  add_profit(c(restore = 8, control = -1, fence = 6)) |>
  add_objective_max_profit(alias = "profit") |>
  set_solver_cbc(gap_limit = 0, verbose = FALSE)
```

```{r solve-economic, eval=requireNamespace("rcbc", quietly = TRUE)}
solution <- solve(economic)
get_actions(solution)
get_objectives(solution)
stopifnot(get_objectives(solution)$profit == 7)
```

The negative profit of control remains included when restoration is selected.
Planning-unit implementation cost is counted once even when several actions
are selected; action costs are counted for each selected action.

These relations also constrain the common feasible set used by weighted-sum,
epsilon-constraint, and AUGMECON methods. Objective definitions, aliases, and
method signatures are unchanged. For example, add a cost objective to the
economic base and choose a profit threshold:

```{r mo-example}
mo <- economic |>
  add_objective_min_cost(alias = "cost", include_pu_cost = FALSE) |>
  set_method_epsilon_constraint(
    primary = "cost", aliases = c("cost", "profit"),
    runs = set_runs_manual(data.frame(eps_profit = 7))
  )
```

```{r solve-mo, eval=requireNamespace("rcbc", quietly = TRUE)}
mo_solution <- solve(mo)
get_objectives(mo_solution)
stopifnot(get_objectives(mo_solution)$profit == 7,
          get_objectives(mo_solution)$cost == 2)
```

Logical coupling does not specify an ecological synergy or antagonism.
Supply joint totals through `add_effects()` to express interactions. Benefit
maximizes signed joint change, loss minimizes final deterioration per unit and
feature, and targets count the reference once; see the
[joint-effects vignette](Joint_effects.html).
