---
title: "Controlling action counts per planning unit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Controlling action counts per planning unit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(multiscape)
```

`add_constraint_action_cardinality()` constrains how many individual actions
are selected **in each planning unit**. It accepts the same `sense`, `actions`,
and optional `name` conventions used by other constraint functions, together
with `count` and the spatial scope `pu`.

| `sense` | Constraint in each scoped unit | Example |
|:--------|:-------------------------------|:--------|
| `"min"` | At least `count` actions | At least one threat-management action |
| `"max"` | At most `count` actions | Capacity for at most four interventions |
| `"equal"` | Exactly `count` actions | Exactly two interventions |

The function adds rules to a new problem object and preserves its input.
Counts are integers, so exact counts do not use a numerical tolerance.

## Define the actions

```{r define-problem}
problem <- create_problem(
  pu = data.frame(id = c(10L, 20L, 30L), cost = 7),
  features = data.frame(id = 1L, name = "woodland"),
  dist_features = data.frame(pu = c(10L, 20L, 30L), feature = 1L, amount = 100)
) |>
  add_actions(
    data.frame(id = c("restore", "control", "fence", "monitor")), cost = 1
  )
```

Without a cardinality rule, the optimizer keeps its default maximum of one
action per unit. Registering action sets does not change that limit and does
not create extra decisions: a set with two members contains two counted
actions, even if one member also belongs to another set.

## Different capacities in different units

```{r different-capacities}
capacities <- problem |>
  add_constraint_action_cardinality(4, "max", pu = 10L, name = "capacity_10") |>
  add_constraint_action_cardinality(2, "max", pu = 20L, name = "capacity_20")
capacities$data$constraints$action_cardinality
```

Unit 10 permits up to four actions, unit 20 up to two, and unit 30 keeps the
default maximum of one. `pu = c(10L, 20L)` would apply a bound separately to
both units, rather than limiting their combined number of actions. `pu = NULL`
applies the rule to all currently registered units.

## Require actions from a subset

```{r subset-minimum}
required <- capacities |>
  add_constraint_action_cardinality(
    1, "min", actions = c("restore", "control"), pu = c(10L, 20L),
    name = "habitat_or_threat"
  )
```

Each of units 10 and 20 must select at least one of `restore` or `control`.
This count includes individual actions, rather than the number of registered
sets. Supply action ids or the existing `actions$action_set` classification
labels; identifiers registered by `add_action_sets()` are not accepted as
selectable actions. Use their individual members instead.

For example, after permitting three total actions per unit, a second rule
can allow at most one from the restoration/control subset:

```{r subset-maximum}
subset_capacity <- problem |>
  add_constraint_action_cardinality(3, "max") |>
  add_constraint_action_cardinality(1, "max", actions = c("restore", "control"))
```

The remaining capacity can be used for fencing and monitoring. A rule that
only counts an action subset does not replace the default total maximum;
the explicit total bound in the first call is therefore necessary here.

## Exact counts and combined bounds

```{r exact-count}
exact <- problem |>
  add_constraint_action_cardinality(2, "equal", pu = 10L) |>
  add_constraint_action_cardinality(0, "equal", pu = 20L)
```

Unit 10 must select exactly two actions; unit 20 cannot select any. An exact
**total** count, like a total maximum, replaces the default one-action limit
in its scoped units. A minimum alone retains that default. To require at least
two actions while allowing up to four, supply both bounds:

```{r minimum-and-maximum}
bounded <- problem |>
  add_constraint_action_cardinality(4, "max", pu = 10L) |>
  add_constraint_action_cardinality(2, "min", pu = 10L)
```

All explicit rules apply together, independently of call order. If a maximum
of four covers units 10 and 20, and a maximum of two covers units 20 and 30,
the effective capacities are four, two, and two:

```{r overlapping-rules}
overlapping <- problem |>
  add_constraint_action_cardinality(4, "max", pu = c(10L, 20L)) |>
  add_constraint_action_cardinality(2, "max", pu = c(20L, 30L))
```

A later call does not overwrite an earlier rule. Reusing exactly the same
planning-unit subset, action subset, and sense is rejected. Constraint names
must also be unique within these rules. `name` labels a constraint for
inspection; objective aliases continue to identify objectives in MO methods.

## Available actions and infeasibility

Only available planning-unit/action pairs are counted. An action that is
locked out or has an invalid cost is excluded as in model compilation.
Locked-in actions remain individual selected decisions and count toward the
bound. A positive minimum or equality larger than the available count in any
scoped unit is rejected. Empty sums are zero: upper bounds and zero bounds
remain valid for units without available actions.

Several individually valid rules can still conflict, or conflict with budgets
and locked decisions. The solver reports that infeasibility. For example,
a minimum of two combined with a maximum of one has no feasible allocation;
adding a minimum does not silently change the upper bound.

## Solve a cost/profit example

This example uses independent economic returns. Four actions in unit 10,
two in unit 20, and one in unit 30 are optimal when every action has positive
profit and no other resource limit applies:

```{r economic-model}
economic <- capacities |>
  add_profit(c(restore = 8, control = 4, fence = 2, monitor = 1)) |>
  add_objective_max_profit(alias = "profit")
```

```{r economic-solution}
if (requireNamespace("rcbc", quietly = TRUE)) {
  solution <- solve(set_solver_cbc(economic, verbose = FALSE, gap_limit = 0))
  selected <- get_actions(solution)
  selected <- selected[selected$selected > 0.5, , drop = FALSE]
  stats::aggregate(selected ~ pu, selected, sum)
  get_objectives(solution)
}
```

The maximum profit is 35: 15 from unit 10, 12 from unit 20, and 8 from unit 30.
Action costs remain associated with each selected action, while planning-unit
costs are charged once for each selected unit. Budgets and net-profit objectives
use those same existing cost definitions. The three MO methods retain the
cardinality constraints while exploring their objectives.

Capacity alone does not specify dependencies or compatibility. Add logical
relations to require companion actions, exclude incompatible actions, or
select an entire group together; see the
[action-relations vignette](Action_relations.html). These relations preserve
the capacities configured here.

## Ecological effects with concurrent actions

Concurrent individual effects are additive unless supplied joint effects
specify a different total; see the [joint-effects vignette](Joint_effects.html).
Benefit maximizes signed change, including negative effects. Loss measures
deterioration after combining all scoped actions within each unit and feature.
Solution summaries and ecological targets count the selected reference once.

```{r additive-ecological-example}
ecological <- create_problem(
  data.frame(id = 10L, cost = 0), data.frame(id = 1L, name = "habitat"),
  data.frame(pu = 10L, feature = 1L, amount = 100)
) |>
  add_actions(data.frame(id = c("restore", "control")), cost = 1) |>
  add_constraint_action_cardinality(2, "equal") |>
  add_effects(data.frame(action = c("restore", "control"),
                        feature = "habitat", effect = c(30, -20))) |>
  add_constraint_targets_absolute(110) |>
  add_objective_max_effect(alias = "benefit")
```

```{r additive-ecological-solution, eval=requireNamespace("rcbc", quietly = TRUE)}
solution <- solve(set_solver_cbc(ecological, gap_limit = 0, verbose = FALSE))
get_objectives(solution)
get_features(solution)
stopifnot(get_objectives(solution)$benefit == 10,
          get_features(solution)$selected_loss == 0,
          get_features(solution)$selected_baseline == 100,
          get_features(solution)$selected_amount_after == 110)
```

The net change is +10, the final loss is zero, and the outcome is 110. The
negative individual contribution is included before the final loss is split.
Existing workflows with at most one action per unit retain that capacity;
`max_benefit()` now includes negative effects in those workflows as well.
