---
title: "Joint effects and interactions between management actions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Joint effects and interactions between management actions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(multiscape)
```

Supply the **total effect of executing a combination** in the same call as its
individual effects. The package preserves those totals and calculates signed
interaction corrections, rather than attributing portions to individual actions.

```{r inputs}
base <- create_problem(
  data.frame(id = 10L, cost = 0), data.frame(id = 1L, name = "habitat"),
  data.frame(pu = 10L, feature = 1L, amount = 100)
) |>
  add_actions(data.frame(id = c("restore", "control")), cost = 1) |>
  add_action_sets(data.frame(set = c("restore_control", "restore_control"),
                             action = c("restore", "control"))) |>
  add_constraint_action_cardinality(2, "max")
effects <- data.frame(action = c("restore", "control", "restore_control"),
                      feature = "habitat", effect = c(30, 20, 70))
problem <- add_effects(base, effects)
problem$data$effect_terms[, c("action", "total_effect", "effect")]
```

The pair correction is `70 - 30 - 20 = 20`. Selecting both yields 70, while
restoration alone yields 30. Registration and effects do not force selection;
cardinality controls capacity and logical relations control permitted groups.

The modern `action` field accepts atomic IDs or registered set IDs. `effect`,
`outcome`, and `relative_change` keep their meanings, sharing the same reference:

```{r equivalent-inputs}
outcomes <- data.frame(action = effects$action, feature = "habitat", outcome = c(130, 120, 170))
changes <- data.frame(action = effects$action, feature = "habitat", relative_change = c(.3, .2, .7))
outcome_problem <- add_effects(base, outcomes)
relative_problem <- add_effects(base, changes)
stopifnot(isTRUE(all.equal(problem$data$effect_terms, outcome_problem$data$effect_terms)),
          isTRUE(all.equal(problem$data$effect_terms, relative_problem$data$effect_terms)))
```

Modern raster lists can likewise include names of registered sets, with the same
layer order, aggregation, reference, and interpretation rules as individual
rasters. Legacy component filtering is rejected for sets. With no `pu`, a joint
row expands only over units where every member has an available pair. Explicit
joint rows with unavailable members are rejected. Later locks or incompatible
constraints can make a supplied combination unreachable without removing its
original observation.

## Preserve inputs and record assumptions

```{r provenance}
problem$data$effects_input
problem$data$joint_effects[, c("action", "effect", "action_outcome")]
problem$data$effects_meta[c("missing_interactions", "missing_individual_effects")]
```

`effects_input` preserves the original table; `effects_original` preserves all
canonical totals; `joint_effects` contains supplied set totals; `effect_terms`
contains signed corrections. `dist_effects` retains individual rows for
compatibility. Different set names cannot provide duplicate membership totals
for one PU/feature. Zero corrections remain preserved as input observations.

Unspecified individual effects and interactions are **assumed zero**. This is
an explicit model assumption, not evidence that an unobserved interaction is
absent. For three actions, subtract individual and supplied pair corrections:

$$\delta_{ABC}=E(ABC)-E(A)-E(B)-E(C)-\delta_{AB}-\delta_{AC}-\delta_{BC}.$$

The rule extends to larger sets. No unprovided subsets are generated automatically.

## Exact activation and model size

For set $S$, compilation adds a continuous auxiliary satisfying:

$$0\le y_{i,S}\le1,\quad y_{i,S}\le x_{i,a}\ (a\in S),\quad
y_{i,S}\ge\sum_{a\in S}x_{i,a}-|S|+1.$$

Binary member decisions determine it exactly, even with negative coefficients
or economic objectives. One PU/membership variable is shared across features.
Only reachable, non-zero corrections need auxiliaries. Availability, known
cardinality limits, and exclusions can eliminate combinations. Inferred negative
final feature amounts are excluded through linear outcome constraints.

Ten actions have 1,013 subsets of size two or greater, but only non-zero,
reachable corrections contribute variables. Sparse interactions limit growth;
solution time still depends on the full model and the number of MO runs.

## Solve with all three MO methods

In this positive-coefficient example, the existing benefit objective includes
individual contributions and the pair correction. Each method selects both
actions, with benefit 70 and action cost 2.

```{r methods}
mo_base <- problem |>
  add_constraint_action_cardinality(1, "min") |>
  add_objective_min_cost(alias = "cost", include_pu_cost = FALSE) |>
  add_objective_max_benefit(alias = "benefit") |>
  set_solver_cbc(gap_limit = 0, verbose = FALSE)
methods <- list(
  weighted = set_method_weighted_sum(mo_base, aliases = c("cost", "benefit"),
    normalize_weights = FALSE, runs = set_runs_manual(data.frame(weight_cost = 1, weight_benefit = 1))),
  epsilon = set_method_epsilon_constraint(mo_base, primary = "cost", aliases = c("cost", "benefit"),
    runs = set_runs_manual(data.frame(eps_benefit = 69))),
  augmecon = set_method_augmecon(mo_base, primary = "cost", aliases = c("cost", "benefit"),
    runs = set_runs_manual(data.frame(eps_benefit = 69)))
)
```

```{r solve-mo, eval=requireNamespace("rcbc", quietly = TRUE)}
solutions <- lapply(methods, solve)
lapply(solutions, get_objectives)
for (solution in solutions) {
  objective <- get_objectives(solution)
  feature <- get_features(solution)
  stopifnot(objective$cost == 2, objective$benefit == 70,
            feature$selected_baseline == 100, feature$selected_amount_after == 170)
}
```

## Negative interactions and current development scope

```{r negative-interaction}
negative_input <- effects
negative_input$effect[negative_input$action == "restore_control"] <- 40
negative <- add_effects(base, negative_input)
negative$data$effect_terms[, c("action", "total_effect", "effect")]
economic <- negative |>
  add_profit(c(restore = 10, control = 9)) |>
  add_objective_max_profit(alias = "profit") |>
  set_solver_cbc(gap_limit = 0, verbose = FALSE)
```

```{r solve-negative, eval=requireNamespace("rcbc", quietly = TRUE)}
solution <- solve(economic)
get_features(solution)
stopifnot(get_features(solution)$selected_net == 40,
          get_features(solution)$selected_benefit == 40,
          get_features(solution)$selected_loss == 0,
          get_features(solution)$selected_amount_after == 140)
```

The correction is -10, but the final change is +40. The summary reports benefit
40 and loss zero, counting the selected reference once. It does not classify
the negative correction as an ecological loss by itself.

Optimizing the final positive/negative split requires the next stage's feature
aggregation formulation. Benefit objectives currently require non-negative
coefficients; loss objectives require non-positive coefficients. Mixed signed
cases raise an explicit error, as do concurrent ecological targets. Internal
SO/MO tests exercise the signed polynomial, including automatic payoff/grid
runs, but no public net-benefit objective is introduced here. Cost/profit models
can use signed joint effects and inspect evaluated ecological summaries.
Individual-only ecological workflows retain their existing behavior and guard.
