---
title: "Repeated calls and alternative problem configurations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Repeated calls and alternative problem configurations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(multiscape)
```

Each constructor has a fixed repeated-call contract. No argument is needed to
choose between accumulation and replacement: existing definitions are never
silently replaced.

| Family | Repeated-call behavior |
|:-------|:-----------------------|
| `add_actions()`, `add_effects()`, `add_profit()` | One complete definition per problem; a second call is an error. |
| `set_solver()` and its backend wrappers | One explicit configuration; a second call is an error. Defaults do not count as configuration. |
| `set_method_*()` | One MO method, including its run design; switching or repeating it is an error. |
| `add_objective_*()` without `alias` | One explicit single-objective definition; a second one is an error. |
| `add_objective_*()` with `alias` | Distinct aliases accumulate; a repeated alias is an error. |
| `add_action_sets()` | Distinct set IDs accumulate; a repeated ID is an error. |
| Targets, area, budget, action cardinality, and logical action relations | Distinct constraint definitions accumulate; repeated identities are errors. |
| Spatial constructors | Distinct relation names accumulate; a repeated name is an error. |
| Planning-unit and action locks | Compatible locks accumulate; identical states are idempotent and contradictory states are errors. |

The legacy `add_benefits()` and `add_losses()` wrappers share the one effects
definition with `add_effects()`. Even a zero-profit specification or an empty
filtered effects table counts as an explicit definition.

## Build a common base

Keep an object before the definition you want to vary. Constructors return an
independent problem, so alternatives can share that common base.

```{r base}
base <- create_problem(
  pu = data.frame(id = 1:3, cost = 1, area = 10),
  features = data.frame(id = 1L, name = "woodland"),
  dist_features = data.frame(pu = 1:3, feature = 1L, amount = 100)
)

actions_base <- base |>
  add_actions(data.frame(id = c("restore", "control", "fence")), cost = 1)

low_effect <- actions_base |>
  add_effects(data.frame(action = "restore", feature = "woodland", effect = 20))

high_effect <- actions_base |>
  add_effects(data.frame(action = "restore", feature = "woodland", effect = 40))

is.null(actions_base$data$dist_effects)
```

Calling `add_effects(low_effect, ...)` again would raise an error. To change
actions, start from `base`; to change effects or profit, start from `actions_base`.

## Compare solver settings and MO methods

Define all solver parameters in one call. Setting a solver does not run it.

```{r alternatives}
objectives_base <- high_effect |>
  add_constraint_targets_absolute(50) |>
  add_objective_min_cost(alias = "cost") |>
  add_objective_max_effect(alias = "benefit")

cbc_problem <- objectives_base |>
  set_solver_cbc(gap_limit = 0, time_limit = 60)

gurobi_problem <- objectives_base |>
  set_solver_gurobi(gap_limit = 0, time_limit = 60)

weighted_problem <- cbc_problem |>
  set_method_weighted_sum(
    aliases = c("cost", "benefit"),
    runs = set_runs_grid(n = 3)
  )

epsilon_problem <- cbc_problem |>
  set_method_epsilon_constraint(
    primary = "cost",
    aliases = c("cost", "benefit"),
    runs = set_runs_manual(data.frame(eps_benefit = c(40, 80)))
  )

augmecon_problem <- cbc_problem |>
  set_method_augmecon(
    primary = "cost",
    aliases = c("cost", "benefit"),
    runs = set_runs_grid(n = 3)
  )

is.null(cbc_problem$data$method)
is.null(objectives_base$data$solve_args)
```

Each method is derived from `cbc_problem`, which has no configured MO method.
To compare run designs within a method, use this same branching pattern.

## Accumulate independent constraints

Constraint identity describes what is constrained, rather than the threshold
or its diagnostic name. Identifier order does not affect identity.
Explicit constraint names must be unique within their family; omitted names
are generated automatically. A different name does not permit a duplicate
constraint identity.

| Constraint | Identity |
|:-----------|:---------|
| Action cardinality | Set of PUs + total/subset action scope + sense |
| Logical action relation | Type + action groups (including required companions) + all/any sense + set of PUs |
| Budget | Action scope + included cost components + sense |
| Area | Action scope + area measure (`area_col`) + sense |
| Target | Feature + action scope, shared by absolute and relative targets |

Requires, excludes, and together share one name namespace. A single required
companion has the same identity under `sense = "all"` and `sense = "any"`.

```{r accumulation}
constrained <- actions_base |>
  add_action_sets(list(restore_control = c("restore", "control"))) |>
  add_action_sets(list(restore_fence = c("restore", "fence"))) |>
  add_constraint_action_cardinality(3, "max", pu = 1) |>
  add_constraint_action_cardinality(2, "max", pu = 2) |>
  add_constraint_budget(20, "max") |>
  add_constraint_budget(10, "max", include_action_cost = FALSE)

get_action_sets(constrained)
constrained$data$constraints$action_cardinality[, c("count", "sense", "name")]
```

Adding another total maximum for PU 1 would be an error, even with a different
count or name. Different scopes may overlap: all accumulated constraints apply
simultaneously. Constraints that are individually valid can still make the model
infeasible; the solver determines joint feasibility.

Absolute and relative targets cannot be used to redefine the same feature and
action scope. Targets for the same feature with different contributing action
subsets can accumulate.

## Accumulate compatible locks

```{r locks}
locked <- actions_base |>
  add_constraint_locked_planning_units(locked_in = 1) |>
  add_constraint_locked_planning_units(locked_out = 3) |>
  add_constraint_locked_actions(locked_out = list(fence = 2)) |>
  add_constraint_locked_actions(locked_out = list(fence = 2))

locked$data$pu[, c("id", "locked_in", "locked_out")]
```

Omitted lock arguments preserve existing states. Repeating the same lock does
not change its meaning. Attempting to lock PU 1 out now would raise an error,
as would locking an action in inside excluded PU 3. This is checked in either
call order. To alter a lock, rebuild from the object before the original lock.

Spatial relations similarly accumulate under distinct names. Reusing a name
raises an error, whether using `add_spatial_relations()` directly or a spatial
constructor such as `add_spatial_boundary()`.

For migration from earlier behavior, move all solver settings into one call,
combine effects into one complete table, and create alternatives from a common
base rather than applying a second configuration to an existing problem.
