---
title: "From reserve selection to multi-objective conservation planning"
output:
  rmarkdown::html_vignette:
    toc: false
    number_sections: true
vignette: >
  %\VignetteIndexEntry{From reserve selection to multi-objective conservation planning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "reserve-design-rwanda-",
  fig.width = 8,
  fig.height = 5.5,
  out.width = "100%",
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  error = FALSE
)

```

# Overview

Systematic conservation planning often begins with a minimum-set question:
which planning units should be included in a reserve network so that all
representation targets are achieved at minimum cost? Spatial configuration can
then be incorporated through a boundary length modifier (BLM), which places
cost and exposed reserve boundary in a single weighted objective. This familiar
formulation is useful when a defensible BLM has already been selected, but it
does not display the full range of attainable compromises between economic and
spatial performance.

This vignette first reproduces that classical reserve-design workflow in
`multiscape`. It then changes the decision question. Instead of selecting a BLM
before optimization, cost and fragmentation are retained as separate objectives
and an epsilon-constraint design is used to reveal their trade-off. The purpose
is not to reproduce the full functionality of dedicated conservation-planning
software. The reserve problem provides a recognizable reference from which to
show how `multiscape` supports explicit multi-objective analysis and comparison
of alternative spatial plans.

The analysis asks:

> Which additional planning units could complement Rwanda's existing protected
> areas to represent at least 30% of every training feature, and what additional
> cost is required to obtain a more spatially compact reserve network?

# The MaPP Rwanda training case

The example adapts the Rwanda tutorial developed for the Marxan Planning
Platform (MaPP) by The Nature Conservancy (TNC 2024). The tutorial contains
2,613 planning units, an existing protected-area layer, a Human Modification
Index-based cost surface, nine mammal distributions, and four terrestrial
ecoregions. Mammal distributions originate from global habitat-suitability
models (Rondinini et al. 2011), ecoregions from the Terrestrial Ecoregions of the World
(Olson et al. 2001), and human modification from Kennedy et al. (2019) and
Theobald et al. (2020).

The MaPP tutorial explicitly describes its narratives as fictional and its
features as an arbitrary subset selected for training. We retain that
interpretation here: the example demonstrates methodology and its solutions
must not be interpreted as conservation recommendations for Rwanda. The source
layers were projected to UTM zone 35S and intersected with the planning units.
A unit is treated as already protected when at least 75% of its area overlaps
the tutorial protected-area layer, following the threshold used in the MaPP
exercise.

The official tutorial archive is distributed by MaPP and can be obtained from
the [MaPP website](https://marxansolutions.org/marxanmapp/). The underlying
spatial layers are not redistributed with `multiscape`. To reproduce the
analysis, download and extract that archive, set `MAPP_RWANDA_DIRECTORY` to the
extracted directory, and run the transparent preparation script supplied in
`data-raw/rwanda/prepare-data.R`. This creates a local working object; it does
not add the tutorial data to the installed package.

```{r load-data}
library(multiscape)
library(dplyr)
library(ggplot2)
library(sf)

Sys.setenv(
  MAPP_RWANDA_DIRECTORY = "path/to/extracted/MaPP-Tutorial-Rwanda"
)

source("data-raw/rwanda/prepare-data.R")
rwanda_reserve <- readRDS(
  "data-raw/rwanda/rwanda-reserve-inputs.rds"
)

```

The processed training case contains 2,613 planning units, of which 201 meet
the tutorial's existing-protection threshold, together with nine mammal
features and four terrestrial ecoregions. The complete list of features and
their spatial summaries can be reproduced with the code above; it is not
stored as package data.

The cost is retained in the scaled units supplied by the tutorial. Larger
values indicate more highly modified planning units and are therefore less
desirable for inclusion in the expanded network. Existing protected units are
locked into every solution and contribute toward the representation targets.

```{r planning-context, fig.width=8, fig.height=5.5, fig.cap="Human-modification cost and existing protected planning units in the MaPP Rwanda training landscape."}
planning_map <- rwanda_reserve$planning_units |>
  mutate(
    protection_status = if_else(
      protected,
      "Existing protected area",
      "Available for selection"
    )
  )

ggplot(planning_map) +
  geom_sf(aes(fill = cost), color = NA) +
  geom_sf(
    data = filter(planning_map, protected),
    fill = NA,
    color = "#1B7837",
    linewidth = 0.25
  ) +
  scale_fill_viridis_c(
    option = "C",
    name = "Human\nmodification"
  ) +
  coord_sf(datum = NA) +
  labs(
    title = "Planning context",
    subtitle = "Green outlines identify units with at least 75% protected-area coverage"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

```{r planning-context-static, echo=FALSE, eval=TRUE, fig.cap="Human-modification cost and existing protected planning units in the MaPP Rwanda training landscape."}
knitr::include_graphics(
  "figures/rwanda/reserve-design-rwanda-planning-context-1.png"
)
```

The feature distributions are spatially heterogeneous. Some are widespread,
whereas others occur in only a small part of the planning region. Amount is
measured as the area of each feature within each planning unit, so targets are
based on represented area rather than simple presence or absence.

```{r feature-distributions, fig.width=9, fig.height=9}
feature_map <- rwanda_reserve$feature_distribution |>
  left_join(rwanda_reserve$features, by = c("feature" = "id")) |>
  group_by(feature) |>
  mutate(relative_amount = amount / max(amount)) |>
  ungroup() |>
  left_join(
    select(rwanda_reserve$planning_units, id, geometry),
    by = c("pu" = "id")
  ) |>
  st_as_sf()

ggplot(feature_map) +
  geom_sf(aes(fill = relative_amount), color = NA) +
  facet_wrap(~name, ncol = 4) +
  scale_fill_viridis_c(
    option = "C",
    limits = c(0, 1),
    name = "Relative\namount"
  ) +
  coord_sf(datum = NA) +
  theme_void(base_size = 9) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 8)
  )
```

```{r feature-distributions-static, echo=FALSE, eval=TRUE, fig.cap="Spatial distributions of the thirteen conservation features. Values are normalized independently within each feature to emphasize spatial pattern rather than differences in total area."}
knitr::include_graphics(
  "figures/rwanda/reserve-design-rwanda-feature-distributions-1.png"
)
```

# Define the reserve problem

Let \(i \in \mathcal{I}\) denote planning units and \(j \in \mathcal{J}\)
denote conservation features. The binary variable \(x_i\) equals one when unit
\(i\) is included in the reserve network. In `multiscape`, this decision is
represented by a single `reserve` action. Because there is only one candidate
action, the action-assignment model reduces to the classical binary
site-selection problem.

The table supplied to `add_effects()` uses `effect_type = "after"`. Its
multiplier of one means that selecting the reserve action represents the
baseline amount of each feature contained in that unit. It does not describe a
100% ecological increase: it describes the amount expected to be represented
after the unit is assigned to the reserve network.

```{r build-problem}
problem <- create_problem(
  pu = rwanda_reserve$planning_units,
  features = rwanda_reserve$features,
  dist_features = rwanda_reserve$feature_distribution,
  cost = "cost",
  pu_id_col = "id"
) |>
  add_actions(
    actions = rwanda_reserve$actions,
    cost = 0
  ) |>
  add_effects(
    effects = rwanda_reserve$effects,
    effect_type = "after"
  ) |>
  add_constraint_targets_relative(0.30) |>
  add_constraint_locked_planning_units(
    locked_in = rwanda_reserve$planning_units$protected
  ) |>
  add_spatial_boundary(
    name = "boundary",
    include_self = TRUE,
    edge_factor = 0.5
  ) |>
  add_objective_min_cost(alias = "cost") |>
  add_objective_min_fragmentation_planning_units(
    relation_name = "boundary",
    alias = "fragmentation"
  )

problem
```

```{r problem-static, echo=FALSE, eval=TRUE, results='asis'}
problem_output <- readLines(
  "results/rwanda/problem-summary.txt",
  warn = FALSE,
  encoding = "UTF-8"
)
cat("```text\n", paste(problem_output, collapse = "\n"), "\n```\n", sep = "")
```

For every feature, the required amount is 30% of its amount across the entire
planning region:

\[
\sum_{i \in \mathcal{I}} r_{ij}x_i
\geq
0.30\sum_{i \in \mathcal{I}}r_{ij},
\qquad j \in \mathcal{J},
\]

where \(r_{ij}\) is the area of feature \(j\) in unit \(i\). Existing protected
units form the set \(\mathcal{P}\) and are fixed through

\[
x_i=1, \qquad i \in \mathcal{P}.
\]

The optimizer therefore identifies additions to the existing network rather
than designing a reserve system from an empty landscape.

# Minimum cost and the boundary length modifier

Without a spatial term, the classical minimum-set objective is

\[
\min C(x)=\sum_i c_i x_i,
\]

where \(c_i\) is the local Human Modification Index-based cost. This formulation
meets the representation requirements as cheaply as possible, but inexpensive
units can form a spatially dispersed network.

A BLM introduces a fragmentation expression \(F(x)\). For adjacent units
\((i,k) \in \mathcal{E}\), let \(b_{ik}\) denote their shared boundary. A
boundary-style expression can be written as

\[
F(x)=\sum_{(i,k)\in\mathcal{E}}b_{ik}|x_i-x_k|,
\]

with the external edge contribution handled by the stored boundary relation.
The BLM formulation is

\[
\min \; C(x)+\lambda F(x),
\]

subject to the same targets and locked-in units. Here \(\lambda\) is the BLM.
It is an exchange coefficient between the units of cost and boundary, not a
percentage of importance.

The manual run design below keeps the coefficient of cost equal to one and
varies only the coefficient of fragmentation. Weight normalization and
objective scaling are deliberately disabled so that the optimized expression
has the traditional BLM interpretation.

```{r define-blm}
blm_values <- c(0, 1e-5, 1e-4, 1e-3, 1e-2, 2e-2, 5e-2, 1e-1, 1)

blm_problem <- problem |>
  set_method_weighted_sum(
    aliases = c("cost", "fragmentation"),
    runs = set_runs_manual(
      data.frame(
        weight_cost = 1,
        weight_fragmentation = blm_values
      )
    ),
    normalize_weights = FALSE,
    objective_scaling = FALSE
  ) |>
  set_solver_gurobi(
    gap_limit = 0,
    verbose = TRUE
  )
```

The analyses shown below were solved with Gurobi to proven optimality
(`gap_limit = 0`) and without a time limit. They are not executed while the vignette is built,
because reproduction requires the external MaPP archive and a licensed solver.
The installed vignette contains only publication figures: neither the spatial
inputs nor the planning-unit selections are distributed with `multiscape`.
After preparing the data locally, the complete analysis can be regenerated by
running `data-raw/build_rwanda_vignette_results.R` from the package source tree.

```{r obtain-solutions}
if (!requireNamespace("gurobi", quietly = TRUE)) {
  stop("Reproducing this analysis requires Gurobi and a valid licence.")
}

blm_solutions <- solve(blm_problem)
```

For transparency, the following excerpt is taken from one of these BLM runs.
It shows the solver version, exact optimality requirement, termination status,
and final zero MIP gap; the remaining runs are summarized in the table below.

```{r gurobi-blm-log-static, echo=FALSE, eval=TRUE, results='asis'}
gurobi_blm_log <- readLines(
  "results/rwanda/gurobi-blm-log-excerpt.txt",
  warn = FALSE
)
cat("```text\n", paste(gurobi_blm_log, collapse = "\n"), "\n```\n", sep = "")
```

The run metadata and objective values should be inspected together. The table
below is the stored output from the seven runs used in this vignette. The
reported gap is the final relative MIP gap returned by Gurobi.

```{r inspect-blm-runs}
blm_run_summary <- get_runs(blm_solutions) |>
  left_join(get_objectives(blm_solutions), by = "solution_id") |>
  mutate(blm = blm_values[run_id]) |>
  select(run_id, solution_id, blm, status, runtime, gap, cost, fragmentation)

blm_run_summary
```

```{r inspect-blm-runs-static, echo=FALSE, eval=TRUE}
blm_run_summary_static <- read.csv(
  "results/rwanda/blm-run-summary.csv",
  check.names = FALSE
)

knitr::kable(
  blm_run_summary_static,
  digits = c(0, 0, 5, NA, 3, 4, 0, 0),
  caption = "Stored run metadata and objective values for the BLM analysis."
)
```

The resulting curve shows what is gained by increasing the BLM. Points that
overlap or are very close indicate ranges of coefficients that induce the same
or nearly the same reserve design. Because the coefficient is scale dependent,
its numerical value should not be transferred unchanged to a landscape with a
different cost surface, projection, resolution, or boundary units.

```{r prepare-blm-results, include=FALSE}
blm_results <- get_runs(blm_solutions) |>
  left_join(get_objectives(blm_solutions), by = "solution_id") |>
  mutate(blm = blm_values[run_id])

minimum_cost <- min(blm_results$cost, na.rm = TRUE)
minimum_cost_fragmentation <- blm_results$fragmentation[
  which.min(blm_results$cost)
]

blm_results <- blm_results |>
  mutate(
    additional_cost = 100 * (cost - minimum_cost) / minimum_cost,
    boundary_km = fragmentation / 1000,
    blm_label = paste0("BLM = ", format(blm, scientific = TRUE))
  )
```

```{r plot-blm, fig.cap="Cost and spatial-configuration outcomes obtained by changing the boundary length modifier. Values are expressed relative to the minimum-cost solution (BLM = 0)."}
ggplot(
  blm_results,
  aes(boundary_km, additional_cost)
) +
  geom_path(color = "grey70", linewidth = 0.5) +
  geom_point(color = "#2C7FB8", size = 2.8) +
  ggrepel::geom_text_repel(
    aes(label = blm_label),
    size = 3,
    min.segment.length = 0,
    seed = 500
  ) +
  scale_y_continuous(labels = function(x) paste0(round(x, 1), "%")) +
  labs(
    x = "Exposed boundary (km)",
    y = "Additional cost relative to the minimum-cost plan",
    title = "BLM calibration",
    subtitle = "Higher BLM values exchange additional cost for a more compact network"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

```{r plot-blm-static, echo=FALSE, eval=TRUE, fig.cap="Cost and spatial-configuration outcomes obtained by changing the boundary length modifier. Boundary is shown in its original spatial units (converted to kilometres), while cost is expressed relative to the minimum-cost solution (BLM = 0)."}
knitr::include_graphics(
  "figures/rwanda/reserve-design-rwanda-plot-blm-1.png"
)
```

# Going beyond a fixed BLM

The BLM formulation is appropriate when an exchange rate between cost and
boundary is already defensible. When that exchange rate is uncertain, keeping
the objectives separate makes the attainable compromises easier to inspect.
For this purpose, cost remains the primary objective and fragmentation is
converted into an explicit upper bound:

\[
\begin{aligned}
\min \quad & C(x) \\
\text{subject to} \quad & F(x) \leq \epsilon_r,
\end{aligned}
\]

in addition to the representation and locked-in constraints. Repeating the
model for different \(\epsilon_r\) values traces the cost of progressively more
compact reserve designs. Epsilon-constraint does not require cost and boundary
to be expressed on a common numerical scale.

```{r define-epsilon}
epsilon_limits <- seq(
  min(blm_results$fragmentation),
  max(blm_results$fragmentation),
  length.out = 8
)

epsilon_problem <- problem |>
  set_method_epsilon_constraint(
    primary = "cost",
    aliases = c("cost", "fragmentation"),
    runs = set_runs_manual(
      data.frame(eps_fragmentation = epsilon_limits)
    )
  ) |>
  set_solver_gurobi(
    gap_limit = 0,
    verbose = TRUE
  )

epsilon_solutions <- solve(epsilon_problem)
```

The excerpt below is retained from the Gurobi output generated while solving
the exact epsilon-constraint analysis. It documents the optimizer, termination
status, objective value, and final optimality gap without requiring the models
to be solved again during website construction.

```{r gurobi-epsilon-log-static, echo=FALSE, eval=TRUE, results='asis'}
gurobi_log <- readLines(
  "results/rwanda/gurobi-epsilon-log-excerpt.txt",
  warn = FALSE
)
cat("```text\n", paste(gurobi_log, collapse = "\n"), "\n```\n", sep = "")
```

The manual epsilon limits span the range of fragmentation values observed during
the BLM calibration. This targets a decision-relevant segment without requiring
a difficult and potentially unrealistic fragmentation-only extreme. In an
applied analysis, the limits could instead be supplied from spatial policy,
implementation requirements, or stakeholder deliberation. Each value is
expressed directly in the units of the stored boundary relation.

```{r prepare-epsilon-results, include=FALSE}
epsilon_results <- get_runs(epsilon_solutions) |>
  left_join(
    get_objectives(epsilon_solutions),
    by = "solution_id"
  )

epsilon_results <- epsilon_results |>
  filter(!is.na(solution_id), is.finite(cost), is.finite(fragmentation)) |>
  mutate(
    additional_cost = 100 * (cost - minimum_cost) / minimum_cost,
    boundary_km = fragmentation / 1000
  )
```

All eight epsilon runs returned feasible solutions within the requested gap.
Here `epsilon_fragmentation` is the imposed upper bound, whereas
`fragmentation` is the boundary value achieved by the corresponding plan.

```{r inspect-epsilon-runs}
epsilon_run_summary <- get_runs(epsilon_solutions) |>
  left_join(get_objectives(epsilon_solutions), by = "solution_id") |>
  mutate(epsilon_fragmentation = epsilon_limits[run_id]) |>
  select(
    run_id,
    solution_id,
    epsilon_fragmentation,
    status,
    runtime,
    gap,
    cost,
    fragmentation
  )

epsilon_run_summary
```

```{r inspect-epsilon-runs-static, echo=FALSE, eval=TRUE}
epsilon_run_summary_static <- read.csv(
  "results/rwanda/epsilon-run-summary.csv",
  check.names = FALSE
)

knitr::kable(
  epsilon_run_summary_static,
  digits = c(0, 0, 0, NA, 3, 4, 0, 0),
  caption = paste(
    "Stored run metadata, epsilon limits, and objective values for the",
    "epsilon-constraint analysis."
  )
)
```

```{r compare-methods, fig.cap="Solutions generated using fixed BLM coefficients and explicit epsilon limits. Epsilon-constraint reveals the attainable cost of progressively reducing exposed boundary."}
comparison_data <- bind_rows(
  blm_results |>
    transmute(
      solution_id,
      additional_cost,
      boundary_km,
      method = "BLM weighted sum"
    ),
  epsilon_results |>
    transmute(
      solution_id,
      additional_cost,
      boundary_km,
      method = "Epsilon-constraint"
    )
)

ggplot(
  comparison_data,
  aes(
    boundary_km,
    additional_cost,
    color = method,
    shape = method
  )
) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_manual(
    values = c(
      "BLM weighted sum" = "#D95F02",
      "Epsilon-constraint" = "#1B9E77"
    )
  ) +
  scale_x_continuous(labels = function(x) paste0(round(x, 1), "%")) +
  scale_y_continuous(labels = function(x) paste0(round(x, 1), "%")) +
  labs(
    x = "Exposed boundary (km)",
    y = "Additional cost relative to the minimum-cost plan",
    color = NULL,
    shape = NULL,
    title = "Two ways to express the same planning tension"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )
```

```{r compare-methods-static, echo=FALSE, eval=TRUE, fig.cap="Solutions generated using fixed BLM coefficients and explicit epsilon limits. Moving toward a lower exposed boundary requires progressively greater cost."}
knitr::include_graphics(
  "figures/rwanda/reserve-design-rwanda-compare-methods-1.png"
)
```

The two methods answer different questions. A BLM solution is conditional on a
specified exchange coefficient. An epsilon solution answers a directly bounded
question such as: what is the least-cost plan if exposed boundary may not exceed
this value? Exploring several bounds supports an a posteriori workflow in which
decision makers inspect the consequences before selecting a plan. A single
policy-derived bound would instead constitute an a priori use of the same
method.

# Compare representative reserve designs

Objective values alone do not reveal whether two plans select the same places.
We therefore retain three representative epsilon solutions: the least-cost
plan, the most compact plan, and an observed compromise closest to the ideal
point after range normalization. The compromise is a descriptive reference,
not an automatically preferred decision.

The maps use `plot_spatial_actions()`, the package function for mapping actions
stored in a `SolutionSet`. Its `nrow` and `ncol` arguments control the facet
arrangement when several solutions are displayed; here the three plans are
placed in two columns. A second layer marks the units that were already
protected, distinguishing inherited protection from additions chosen by the
optimizer.

```{r select-plans, include=FALSE}
cost_range <- diff(range(epsilon_results$cost))
fragmentation_range <- diff(range(epsilon_results$fragmentation))

epsilon_ranked <- epsilon_results |>
  mutate(
    normalized_cost = if (cost_range > 0) {
      (cost - min(cost)) / cost_range
    } else {
      0
    },
    normalized_fragmentation = if (fragmentation_range > 0) {
      (fragmentation - min(fragmentation)) / fragmentation_range
    } else {
      0
    },
    distance_to_ideal = sqrt(
      normalized_cost^2 + normalized_fragmentation^2
    )
  )

representative_ids <- c(
  "Minimum cost" = epsilon_ranked$solution_id[which.min(epsilon_ranked$cost)],
  "Observed compromise" = epsilon_ranked$solution_id[
    which.min(epsilon_ranked$distance_to_ideal)
  ],
  "Minimum fragmentation" = epsilon_ranked$solution_id[
    which.min(epsilon_ranked$fragmentation)
  ]
)

representative_ids <- representative_ids[!duplicated(representative_ids)]

action_results <- get_actions(epsilon_solutions)

selected_units <- action_results |>
  filter(
    solution_id %in% unname(representative_ids),
    selected == 1
  ) |>
  mutate(
    plan = names(representative_ids)[
      match(solution_id, representative_ids)
    ]
  ) |>
  select(pu, solution_id, plan)

representative_maps <- bind_rows(lapply(
  names(representative_ids),
  function(plan_name) {
    rwanda_reserve$planning_units |>
      select(id, protected) |>
      mutate(plan = plan_name)
  }
)) |>
  left_join(selected_units, by = c("id" = "pu", "plan" = "plan")) |>
  mutate(
    status = case_when(
      protected ~ "Existing protected area",
      !is.na(solution_id) ~ "Additional selected unit",
      TRUE ~ "Not selected"
    ),
    plan = factor(plan, levels = names(representative_ids))
  )
```

```{r representative-maps, fig.width=9, fig.height=4.8, fig.cap="Representative reserve designs from the epsilon-constraint frontier. Green units were already protected; orange units are additional selections made under different spatial limits."}
action_plot <- plot_spatial_actions(
  epsilon_solutions,
  solutions = unname(representative_ids),
  layout = "single",
  ncol = 2,
  fill_values = c(reserve = "#F39C12"),
  base_fill = "#F2F2F2",
  base_alpha = 1
)

action_plot +
  facet_wrap(
    ~solution_id,
    ncol = 2,
    labeller = as_labeller(
      setNames(names(representative_ids), representative_ids)
    )
  ) +
  geom_sf(
    data = filter(rwanda_reserve$planning_units, protected),
    fill = "#238B45",
    color = NA
  ) +
  labs(title = "Alternative reserve designs")
```

```{r representative-maps-static, echo=FALSE, eval=TRUE, fig.cap="Representative reserve designs from the epsilon-constraint frontier. Existing protected units occur in every plan; orange units are additions selected to meet the 30% targets under different spatial limits."}
knitr::include_graphics(
  "figures/rwanda/reserve-design-rwanda-representative-maps-1.png"
)
```

The maps distinguish economic and spatial trade-offs from changes in location.
Two plans can have similar objective values while prescribing different
additions to the protected-area network. Conversely, some units may occur
consistently because they contain restricted features, complement the existing
network, or provide an efficient spatial connection.

Selection frequency summarizes this stability across all epsilon solutions.
It should not be interpreted as a probability of protection: it is the number
or proportion of analyzed planning scenarios in which a unit was selected.

```{r selection-frequency, fig.width=7.5, fig.height=5.2, fig.cap="Selection frequency across the epsilon-constraint solutions. Existing protected units are shown separately because they were locked into every plan."}
frequency_data <- action_results |>
  filter(selected == 1) |>
  count(pu, name = "selected_runs") |>
  mutate(frequency = selected_runs / nrow(epsilon_results))

frequency_map <- rwanda_reserve$planning_units |>
  left_join(frequency_data, by = c("id" = "pu")) |>
  mutate(
    frequency = coalesce(frequency, 0),
    frequency_available = if_else(protected, NA_real_, frequency)
  )

ggplot(frequency_map) +
  geom_sf(aes(fill = frequency_available), color = NA) +
  geom_sf(
    data = filter(frequency_map, protected),
    fill = "#238B45",
    color = NA
  ) +
  scale_fill_viridis_c(
    option = "B",
    limits = c(0, 1),
    na.value = "#238B45",
    name = "Selection\nfrequency"
  ) +
  coord_sf(datum = NA) +
  labs(
    title = "Spatial stability across the frontier",
    subtitle = "Green units were already protected and locked into every solution"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

```{r selection-frequency-static, echo=FALSE, eval=TRUE, fig.cap="Selection frequency across the epsilon-constraint solutions. Existing protected units are shown separately because they were locked into every plan."}
knitr::include_graphics(
  "figures/rwanda/reserve-design-rwanda-selection-frequency-1.png"
)
```

# Interpretation

The minimum-set, BLM, and epsilon-constraint formulations are not competing
definitions of the reserve problem. They represent progressively different
preference statements. Minimum cost assumes that the representation targets
and existing protected network are sufficient to define feasibility. The BLM
adds a pre-specified exchange rate between economic and spatial performance.
Epsilon-constraint instead treats spatial configuration as an explicit limit
and can be repeated to characterize the consequences of alternative limits.

This distinction illustrates the broader role of `multiscape`. A familiar
reserve-selection model can be retained when it matches the decision context,
while its constituent criteria can also be registered as separate objectives
when trade-offs should remain visible. The resulting frontier does not choose a
reserve design on behalf of decision makers. It organizes a set of feasible,
interpretable alternatives that all satisfy the same 30% representation
requirements.

# References




```{r references, results='asis', echo=FALSE, eval=TRUE}
cat(paste0(
  "TNC. 2024. *The Marxan Planning Platform: Tutorial (Rwanda).* ",
  "Global Science, The Nature Conservancy, Arlington, Virginia, USA. ",
  "<https://marxanplanning.org>.\n\n",
  "Kennedy, C. M., et al. 2019. Managing the middle: a shift in conservation ",
  "priorities based on the global human modification gradient. ",
  "*Global Change Biology* 25: 811-826. ",
  "<https://doi.org/10.1111/gcb.14549>.\n\n",
  "Olson, D. M., et al. 2001. Terrestrial Ecoregions of the World: A New Map ",
  "of Life on Earth. *BioScience* 51: 933-938. ",
  "<https://doi.org/10.1641/0006-3568(2001)051%5B0933:TEOTWA%5D2.0.CO;2>.\n\n",
  "Rondinini, C., et al. 2011. Global habitat suitability models of ",
  "terrestrial mammals. *Philosophical Transactions of the Royal Society B* ",
  "366: 2633-2641. <https://doi.org/10.1098/rstb.2011.0113>.\n\n",
  "Theobald, D. M., et al. 2020. Earth transformed: detailed mapping of ",
  "global human modification from 1990 to 2017. ",
  "*Earth System Science Data* 12: 1953-1972. ",
  "<https://doi.org/10.5194/essd-12-1953-2020>."
))
```
