---
title: "Simulated multi-objective workflow"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```

## Getting started

### The planning problem

The following example represents a stylised **multi-objective multi-use spatial planning
problem**. The landscape is divided into 64 planning units and contains two
ecological features: woodland and riparian habitat. Each planning unit can
remain unmanaged or be assigned to one of two mutually exclusive management
actions: **protection** or **restoration**.

For every planning unit \(i\) and action \(a\), the model defines a binary
decision variable \(x_{ia}\). The variable equals 1 when action \(a\) is
assigned to planning unit \(i\), and 0 otherwise. Because the actions are
mutually exclusive, at most one action can be selected in each planning unit.
A planning unit may also remain without intervention.

The analysis asks:

> How does the spatial allocation of protection and restoration change as
> progressively larger implementation budgets become available, and how much
> additional ecological benefit can be obtained?

For each planning unit, `amount` is a synthetic **relative feature amount**.
Baseline values range from 0 to 1: a value of 0 represents no local amount of
the feature, whereas 1 is the largest baseline amount represented in the
simulated landscape. Intermediate values are continuous relative amounts, not
probabilities of occurrence or binary presence--absence observations.
Woodland forms a north-western hotspot, whereas riparian habitat follows a
diagonal corridor.

The 0--1 scale applies to the baseline values only. Because these are relative
indices rather than probabilities or proportions, action-induced final amounts
may exceed 1. For example, doubling a baseline amount of 0.6 produces a final
relative amount of 1.2.

All inputs are included with `multiscape`:

```{r example-data}
# load packages
library(multiscape)
library(dplyr)

# Load a complete simulated planning problem.
example_data <- load_sim_multiaction()
```

The object contains the following linked tables:

- `planning_units`: planning-unit identifiers, geometries, and baseline costs;
- `features`: identifiers and names of the ecological features;
- `dist_features`: baseline feature amounts by planning unit;
- `actions`: the available management actions;
- `action_costs`: the cost of each feasible planning-unit--action combination;
- `effects`: the expected effect of each action on each feature; and
- `effect_assumptions`: the coefficients used to generate the simulated
  effects.

The simulated baseline feature amounts are shown below.

```{r example-features, echo=FALSE, fig.width=8.5, fig.height=4.8}
feature_map <- sf::st_as_sf(merge(
  example_data$planning_units,
  merge(
    example_data$dist_features,
    example_data$features,
    by.x = "feature",
    by.y = "id"
  ),
  by.x = "id",
  by.y = "pu"
))
feature_map$feature_label <- factor(
  feature_map$name,
  levels = c("woodland", "riparian"),
  labels = c("Woodland", "Riparian habitat")
)

ggplot2::ggplot(feature_map) +
  ggplot2::geom_sf(
    ggplot2::aes(fill = amount),
    colour = "white",
    linewidth = 0.18
  ) +
  ggplot2::facet_wrap(~feature_label, nrow = 1) +
  ggplot2::scale_fill_gradientn(
    colours = c("#F7FCF5", "#C7E9C0", "#74C476", "#238B45", "#00441B"),
    limits = c(0, 1),
    breaks = c(0, 0.25, 0.50, 0.75, 1),
    name = "Relative feature amount"
  ) +
  ggplot2::coord_sf(expand = FALSE) +
  ggplot2::guides(
    fill = ggplot2::guide_colourbar(
      title.position = "top",
      title.hjust = 0.5,
      barwidth = grid::unit(11, "cm"),
      barheight = grid::unit(0.45, "cm")
    )
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text = ggplot2::element_blank(),
    axis.title = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(face = "bold", size = 11),
    legend.position = "bottom",
    legend.margin = ggplot2::margin(t = 4),
    plot.margin = ggplot2::margin(5, 8, 5, 8)
  )
```

### Stage 1: Create the base problem

[`create_problem()`](https://josesalgr.github.io/multiscape/reference/create_problem.html)
establishes the spatial domain and the ecological features
being planned. Here, `planning_units` contains the 64 geometries, `features`
defines the feature catalogue, and `dist_features` links every planning unit to
its woodland and riparian amounts. The planning-unit `cost` column is retained
in the problem, although the objective used below includes only
action-specific implementation costs.

```{r example-stage-1}
# Initialise the problem using planning-unit geometries and baseline feature
# amounts.
problem <- create_problem(
  pu = example_data$planning_units,
  features = example_data$features,
  dist_features = example_data$dist_features,
  cost = "cost"
)
```

### Stage 2: Define actions and their effects

[`add_actions()`](https://josesalgr.github.io/multiscape/reference/add_actions.html)
registers protection and restoration as alternative management
uses, associates each action with its local implementation cost, and defines
the feasible planning-unit--action combinations. Both actions are available
throughout this landscape. The model permits at most one selected action per
planning unit, while leaving a unit unmanaged remains feasible.

Effects describe changes relative to the reference scenario in `dist_features`.
[`add_effects()`](https://josesalgr.github.io/multiscape/reference/add_effects.html)
accepts exactly one of three columns:

- `effect`: signed absolute change;
- `outcome`: expected amount under the action;
- `relative_change`: proportional change (`0.25` means +25%).

With a reference amount of 100, `effect = 30`, `outcome = 130`, and
`relative_change = 0.30` all describe the same result. A positive effect means
an increase; whether that is desirable depends on the feature and objective.
The reference may represent current conditions, a future without intervention,
or existing management. Outcomes and references must share units and horizon.

This example supplies relative changes by action and feature. The package
expands them over feasible planning units. A reference of 0.6 and
`relative_change = 1` produce an effect of 0.6 and an outcome of 1.2.
Legacy `effect_type`, `multiplier`, and `delta` inputs remain supported with
a lifecycle deprecation warning.

```{r example-stage-2}
# Inspect the assumptions used to generate their
# ecological effects, and the first planning-unit--action--feature records.
example_data$effect_assumptions

problem <- problem |>
  add_actions(
    actions = example_data$actions,
    cost = example_data$action_costs
  ) |>
  add_effects(
    effects = example_data$effect_assumptions
  )
```

### Repeated calls and alternative configurations

Actions, effects, profit, solver settings, an MO method, and an unaliased
single objective can each be defined **once per problem**. A second call raises
an error, even with identical inputs. This also applies to solver and effects
wrappers. Put all solver settings and the complete effects table in their first
calls. Defaults do not count as an explicit configuration.

Distinct action sets, objective aliases, spatial relation names, and constraint
definitions accumulate. Duplicate definitions raise an error. Absolute and
relative targets share the same feature/action-scope identity. Locks accumulate
compatible states; identical repeats are allowed and contradictions are rejected.

To compare configurations, keep a common problem before the setting to vary:

```{r independent-solver-settings}
cbc_problem <- problem |> set_solver_cbc(gap_limit = 0, time_limit = 60)
gurobi_problem <- problem |> set_solver_gurobi(gap_limit = 0, time_limit = 60)
```

Both alternatives have their own solver configuration. Apply the same pattern
when comparing effect scenarios, MO methods, or run designs. See the
[repeated-call contract](https://josesalgr.github.io/multiscape/articles/Repeated_calls.html)
for identities, accumulation rules, and migration examples.

### Optional: register action sets

[`add_action_sets()`](https://josesalgr.github.io/multiscape/reference/add_action_sets.html)
records named combinations of existing actions. It accepts a named list or a
long table with columns `set` and `action`. One action may belong to several
sets. The following independent example defines restoration with threat control
and restoration with fencing:

```{r action-set-definitions}
actions_problem <- create_problem(
  pu = data.frame(id = 1:2, cost = 1),
  features = data.frame(id = 1L, name = "woodland"),
  dist_features = data.frame(pu = 1:2, feature = 1L, amount = 10)
) |>
  add_actions(
    actions = data.frame(id = c("restore", "control", "fence")),
    cost = 1
  )

memberships <- data.frame(
  set = c("restore_control", "restore_control", "restore_fence", "restore_fence"),
  action = c("restore", "control", "restore", "fence")
)
sets_problem <- add_action_sets(actions_problem, memberships)
get_action_sets(sets_problem)

# The named-list input produces the same definitions.
list_problem <- add_action_sets(actions_problem, list(
  restore_control = c("restore", "control"),
  restore_fence = c("restore", "fence")
))
identical(get_action_sets(sets_problem), get_action_sets(list_problem))
```

Registering sets leaves the individual actions and their feasible pairs intact.
It does not require joint selection, introduce interactions, or enable multiple
selected actions per unit. Modern `add_effects()` accepts set identifiers for
total joint effects. Objectives and constraints still receive individual action ids.
The [action-set vignette](https://josesalgr.github.io/multiscape/articles/Action_sets.html)
explains validation and adding new sets across calls.

### Optional: configure action counts per unit

[`add_constraint_action_cardinality()`](https://josesalgr.github.io/multiscape/reference/add_constraint_action_cardinality.html)
sets a minimum, maximum, or exact number of individual actions in each specified
planning unit. Multiple calls can define different capacities or constrain
particular action subsets. This independent example allows up to four actions
in unit 10 and two in unit 20; unit 30 keeps the default maximum of one:

```{r action-cardinality-definitions}
count_problem <- create_problem(
  pu = data.frame(id = c(10L, 20L, 30L), cost = 1),
  features = data.frame(id = 1L, name = "woodland"),
  dist_features = data.frame(pu = c(10L, 20L, 30L), feature = 1L, amount = 10)
) |>
  add_actions(
    data.frame(id = c("restore", "control", "fence", "monitor")), cost = 1
  ) |>
  add_constraint_action_cardinality(4, "max", pu = 10L) |>
  add_constraint_action_cardinality(2, "max", pu = 20L) |>
  add_constraint_action_cardinality(
    1, "min", actions = c("restore", "control"), pu = c(10L, 20L)
  )
count_problem$data$constraints$action_cardinality
```

An explicit total maximum or equality (`actions = NULL`) replaces the implicit
one-action maximum only in the units it covers. Subset rules and minima alone
retain that default. Overlapping explicit rules must all be satisfied;
later calls do not overwrite earlier ones. Registering a set adds no counted
decision. The optional `name` labels a constraint independently of objective aliases.

Concurrent actions support cost/profit workflows and registered joint effects.
Mixed-sign benefit/loss objectives and concurrent ecological targets remain
protected until final feature aggregation is implemented. The main ecological
example below retains its one-action maximum. See the
[action-cardinality vignette](https://josesalgr.github.io/multiscape/articles/Action_cardinality.html)
for exact counts, subset rules, and a solved economic example.

### Optional: define logical relations between actions

Logical relations constrain **which actions can be selected together in each
unit**. They complement cardinality, which controls how many actions can be
selected. Distinct calls accumulate; duplicate definitions or names are errors.

```{r action-relation-definitions}
relations_problem <- actions_problem |>
  add_constraint_action_cardinality(3, "max") |>
  # Restoration needs threat control in unit 1.
  add_constraint_action_requires("restore", "control", pu = 1L) |>
  # In unit 2, either control or fencing is sufficient.
  add_constraint_action_requires(
    "restore", c("control", "fence"), sense = "any", pu = 2L
  ) |>
  # Control and fencing are alternative interventions in both units.
  add_constraint_action_excludes(c("control", "fence"))

# A separate alternative implements restoration and control together or neither.
together_problem <- actions_problem |>
  add_constraint_action_cardinality(2, "max") |>
  add_constraint_action_together(c("restore", "control"))
relations_problem$data$constraints$action_relations
```

`requires` is directional: selecting control does not force restoration.
`together` requires all group members or none; `excludes` permits at most one
member, including zero. An unavailable companion prevents an all dependency;
an any dependency can use the remaining companions. An unavailable together
member prevents the whole group. These rules
apply to the feasible set shared by all optimization methods. They introduce
no ecological interaction coefficients. See the
[action-relations vignette](https://josesalgr.github.io/multiscape/articles/Action_relations.html)
for solved examples, set members, scope, and compatibility with cardinality.

### Optional: supply joint action effects

Supply total joint effects alongside individual effects in **one call** to
`add_effects()`. The package preserves supplied totals and computes signed
interaction corrections. The following independent example has a reference
amount of 100 and a joint change of 70:

```{r joint-effect-definitions}
joint_problem <- create_problem(
  data.frame(id = 10L, cost = 0), data.frame(id = 1L, name = "habitat"),
  data.frame(pu = 10L, feature = 1L, amount = 100)
) |>
  add_actions(data.frame(id = c("restore", "control")), cost = 1) |>
  add_action_sets(list(restore_control = c("restore", "control"))) |>
  add_constraint_action_cardinality(2, "max") |>
  add_effects(data.frame(
    action = c("restore", "control", "restore_control"),
    feature = "habitat", effect = c(30, 20, 70)
  ))
joint_problem$data$effect_terms[, c("action", "total_effect", "effect")]
```

The pair correction is `70 - 30 - 20 = 20`. An exact AND auxiliary contributes
it only when both members are selected, and is shared across features. Only
non-zero, reachable corrections need auxiliaries. `outcome`, `relative_change`,
and modern raster lists can also describe registered sets. Missing individual
effects and interactions are explicitly assumed zero, as recorded in
`effects_meta`. These assumptions do not establish empirical absence of
interactions.

The [joint-effects vignette](https://josesalgr.github.io/multiscape/articles/Joint_effects.html)
includes solved examples with all three MO methods and negative interactions.
Public benefit/loss objectives currently require sign-compatible coefficients;
mixed cases and concurrent ecological targets await the next aggregation stage.
Joint solution summaries already evaluate final signed effects and count the
selected reference once.

### Stage 3: Define the constraints

[`add_constraint_targets_relative()`](https://josesalgr.github.io/multiscape/reference/add_constraint_targets_relative.html)
requires the selected actions to generate gains equivalent
to at least 10% of the total baseline amount of **each feature separately**.
Targets restrict the feasible set; they are not additional objectives. Other
applications could introduce budgets, area requirements, locked decisions, or
spatial constraints at this stage.

```{r example-stage-3, message=TRUE}
# Require action-induced gains of at least 10% of the total baseline amount of
# woodland and at least 10% of the total baseline amount of riparian habitat.
problem <- problem |>
  add_constraint_targets_relative(0.10)
```

### Stage 4: Define the objectives

Objectives are registered independently so they can later be combined using
different multi-objective methods.
[`add_objective_min_cost()`](https://josesalgr.github.io/multiscape/reference/add_objective_min_cost.html)
minimises the total
implementation cost of the selected actions. Planning-unit costs are excluded
because the example defines economic expenditure through the action-specific
cost table.
[`add_objective_max_effect()`](https://josesalgr.github.io/multiscape/reference/add_objective_max_benefit.html)
maximises total ecological benefit, calculated
as the sum of action-induced gains across selected planning units and features.

Because woodland and riparian effects use the same relative scale and no
feature-specific weights are supplied, the example gives both features equal
weight in the benefit objective. See the
[`add_objective_max_effect()` reference](https://josesalgr.github.io/multiscape/reference/add_objective_max_benefit.html)
for the complete definition of the objective, its available arguments, and
how feature contributions are aggregated.

```{r example-stage-4, message=TRUE}
problem <- problem |>
  # Minimise action-specific implementation costs. Planning-unit costs are
  # excluded because expenditure is represented by the action-cost table.
  add_objective_min_cost(
    alias = "cost",
    include_pu_cost = FALSE,
    include_action_cost = TRUE
  ) |>
  # Maximise the sum of action-induced ecological gains across both features.
  add_objective_max_effect(alias = "benefit")
```

The summary confirms that the spatial inputs, actions, effects, constraints,
and two objectives are present. It also shows that the multi-objective method
and solver have not yet been selected.

````{r example-stage-4-1, message=TRUE}
# Print the complete formulation before selecting a multi-objective method
# and optimisation solver.
problem
````

### Configure the multi-objective method

The registered objectives define what the plans should achieve; the
multi-objective method determines how the trade-off between them is explored.
Here,
[`set_method_epsilon_constraint()`](https://josesalgr.github.io/multiscape/reference/set_method_epsilon_constraint.html)
treats ecological benefit as the primary
objective and evaluates six increasingly permissive limits on total cost. For
each cost limit, the model identifies the feasible plan with the greatest
ecological benefit.

With `lexicographic = TRUE`, solutions tied on the primary benefit objective are
refined by minimising cost. This prevents the method from returning an
unnecessarily expensive plan when a less costly plan attains the same benefit.
The six limits are generated by
[`set_runs_grid()`](https://josesalgr.github.io/multiscape/reference/set_runs_grid.html).
The example uses Gurobi through
[`set_solver_gurobi()`](https://josesalgr.github.io/multiscape/reference/set_solver_gurobi.html),
which requires the Gurobi R package and a valid licence.

```{r example-method}
problem <- problem |>
  # Maximise benefit under six alternative limits on total cost.
  # Lexicographic refinement selects the least-cost plan among solutions tied
  # on the primary objective.
  set_method_epsilon_constraint(
    primary = "benefit",
    runs = set_runs_grid(6),
    lexicographic = TRUE
  ) |>
  set_solver_gurobi(gap_limit = 0)
```

### Solve the problem

[`solve()`](https://josesalgr.github.io/multiscape/reference/solve.html) returns a
`SolutionSet` containing the solver status, objective
values, selected action assignments, and run-level information for every
epsilon-constraint problem.
[`get_objectives()`](https://josesalgr.github.io/multiscape/reference/get_objectives.html)
extracts the resulting objective
values, with one row per stored solution and one column per objective.

```{r example-solve}
solutions <- solve(problem)
get_objectives(solutions, format = "wide")
```

### Link run configurations to stored solutions

A multi-objective method evaluates a sequence of optimisation configurations.
[`get_runs()`](https://josesalgr.github.io/multiscape/reference/get_runs.html)
returns the registry of these configurations together with their solver
outcomes and, when available, the identifier of the solution they produced.

```{r example-runs}
runs <- get_runs(solutions)
runs
```

Each row records one attempted run configuration. `run_id` identifies the
configuration, including its epsilon limit and solver outcome. A run is not
itself a solution: an infeasible configuration, a run that terminates without a
feasible solution, or a run affected by a solver error may have no associated
`solution_id`. When a run does produce a solution that is retained in the
`SolutionSet`, `solution_id` identifies that stored solution for subsequent
extraction, comparison, and mapping. In this example, all configured runs
produce stored solutions.

The treatment of infeasible runs, runs without a solution, and unexpected
errors can be controlled with
[`set_runs_control()`](https://josesalgr.github.io/multiscape/reference/set_runs_control.html).
Depending on these settings, solving can stop when a failed configuration is
encountered or retain the failed run in the run history without an associated
solution.

This distinction is important when reading the figures below. With
`label_runs = TRUE`, `plot_tradeoff()` labels points using `run_id`, whereas
functions that extract or map stored plans generally use `solution_id`.
`get_runs()` provides the link between the run configurations and the solutions
that were successfully stored.

## Interpret the solutions

### Objective space: performance and trade-offs

Each point below represents a stored spatial solution generated by one
epsilon-constraint run. The plot is produced with
[`plot_tradeoff()`](https://josesalgr.github.io/multiscape/reference/plot_tradeoff.html).
With lexicographic refinement, the returned plans are
intended to lie on the **observed efficient frontier**. This is the frontier
represented by the six runs evaluated here, not necessarily every attainable
trade-off in the complete discrete solution space.

The horizontal axis shows the cost actually incurred by each solution, which
may be lower than the upper cost limit imposed in its run. The vertical axis
shows total ecological benefit, obtained by summing the action-induced gains
across both features and all selected planning units. Moving from left to right
therefore reveals how the maximum observed ecological benefit changes as more
expensive plans become available.

```{r example-tradeoff, fig.width=6.5, fig.height=4.5, fig.keep="last"}
plot_tradeoff(
  solutions,
  objectives = c("cost", "benefit"),
  connect = TRUE,
  label_runs = TRUE
) +
  ggplot2::labs(
    x = "Total implementation cost (cost units)",
    y = "Total ecological gain (relative-amount units)"
  )
```

The curve rises rapidly at first and becomes flatter near its high-benefit end.
In this region, substantial increases in implementation cost produce only small
additional ecological gains. The result illustrates declining marginal returns
and helps identify where further expenditure has comparatively little effect.

### Identify a compromise solution

A multi-objective analysis should consider both **objective space**, which
describes how well each plan performs, and **decision space**, which describes
where actions are allocated.
[`frontier_knee()`](https://josesalgr.github.io/multiscape/reference/frontier_knee.html)
identifies an empirical compromise on the observed frontier:
a solution near the point where additional ecological gains begin to require
comparatively large increases in cost.

```{r example-knee}
knee <- frontier_knee(
  solutions,
  objectives = c("cost", "benefit")
)
knee |> dplyr::select(solution_id, cost, benefit, knee_score, method)
```

The knee is a decision aid rather than a universally optimal answer. Its
location depends on the solutions included in the observed frontier, the
normalisation of the objectives, and the geometric criterion used to calculate
the knee score. It should therefore be interpreted alongside policy
preferences and the spatial characteristics of the selected plan.

[`frontier_extremes()`](https://josesalgr.github.io/multiscape/reference/frontier_extremes.html) and
[`frontier_distances()`](https://josesalgr.github.io/multiscape/reference/frontier_distances.html) provide complementary views of
the observed objective space. They identify the objective-wise extremes and
calculate each solution's normalised distance from the observed ideal and
nadir points.

```{r example-frontier-diagnostics}
frontier_extremes(
  solutions,
  objectives = c("cost", "benefit"),
  ties = "first"
)

frontier_distances(
  solutions,
  objectives = c("cost", "benefit")
)
```

### Decision space: spatial prescriptions

Plans that are close in objective space can still prescribe different actions
in different locations.
[`plot_spatial_actions()`](https://josesalgr.github.io/multiscape/reference/plot_spatial_actions.html)
complements the trade-off plot
by showing how protection and restoration are reallocated as the cost limit is
relaxed. Planning units shown only through the base layer receive no management
action in the corresponding solution. In the call below, `solutions = 1:6`
selects **solution IDs**, not run IDs; the preceding run registry provides the
link between the two identifiers.

```{r example-maps, fig.width=10, fig.height=6.5}
plot_spatial_actions(
  solutions,
  solutions = 1:6,
  fill_values = c(
    protect = "#2E7D32",
    restore = "#E69F00"
  ),
  base_alpha = 0.12
)
```

The sequence of maps should be interpreted together with the baseline feature
patterns. Protection and restoration may expand into new planning units as
larger budgets become available, but actions can also be substituted or
relocated when a different spatial combination produces greater benefit under
the active cost limit.

### Compare spatial similarity

Similar objective values do not necessarily imply similar spatial
prescriptions.
[`selection_similarity()`](https://josesalgr.github.io/multiscape/reference/selection_similarity.html)
compares the sets of selected
planning-unit--action pairs for every pair of solutions. Jaccard similarity is
useful here because it focuses on jointly selected assignments and does not
inflate similarity through planning units that remain unmanaged in both plans.

A value of 1 indicates identical sets of selected assignments, whereas 0
indicates that the two solutions share no selected planning-unit--action pair.
Rows and columns correspond to solution identifiers, and the diagonal is always
1 because every solution is identical to itself.

```{r example-similarity}
selection_similarity(
  solutions,
  metric = "jaccard",
  format = "matrix"
)
```

### Identify recurrent action assignments

[`selection_frequency()`](https://josesalgr.github.io/multiscape/reference/selection_frequency.html)
calculates how often each planning-unit--action pair is
selected across the solution set. In this example, frequency is the proportion
of the six solutions containing a particular assignment. High frequency
indicates recurrence across the explored trade-offs, but it should not, on its
own, be interpreted as ecological irreplaceability.

```{r example-frequency}
action_frequency <- selection_frequency(solutions)
head(
  action_frequency[order(-action_frequency$frequency), ],
  10
)
```

### Link objective performance to spatial change

Objective-space proximity does not necessarily imply similar spatial
prescriptions.
[`frontier_neighbors()`](https://josesalgr.github.io/multiscape/reference/frontier_neighbors.html)
identifies local neighbours along the observed objective-space trade-off. Here,
solutions are oriented towards increasing ecological benefit, matching the
progression from lower- to higher-budget plans used above.
[`linkage_distances()`](https://josesalgr.github.io/multiscape/reference/linkage_distances.html)
then compares those same solution pairs in objective and decision space.

```{r example-linkage}
neighbors <- frontier_neighbors(
  solutions,
  objectives = c("benefit", "cost")
)

linkage <- linkage_distances(
  solutions,
  objectives = c("cost", "benefit"),
  pairs = neighbors,
  decision_metric = "jaccard"
)

linkage |>
  dplyr::select(
    from_solution,
    to_solution,
    objective_distance,
    decision_distance,
    changed_planning_units
  )
```

The two distances describe different aspects of the same comparison:
`objective_distance` measures separation in normalised objective space,
whereas `decision_distance` measures dissimilarity between spatial action
allocations. Keeping them separate makes it possible to identify similar
levels of performance that require substantially different spatial plans.

[`linkage_transition()`](https://josesalgr.github.io/multiscape/reference/linkage_transition.html)
provides a more detailed view of a selected comparison. Here, we examine the
transition from solution 3 to solution 4.

```{r example-transition}
transition <- linkage_transition(
  solutions,
  from = 3,
  to = 4,
  objectives = c("cost", "benefit")
)

transition
```

The printed summary reports objective and decision distances, the number of
planning units whose state changes, and the corresponding changes in each
objective. Detailed planning-unit, action, and state-transition tables remain
available within the returned object.

[`linkage_turnover()`](https://josesalgr.github.io/multiscape/reference/linkage_turnover.html)
extends these local comparisons by relating decision-space turnover to the
associated objective-space change. Its `reconfiguration_rate` highlights local
transitions that require comparatively large spatial changes for a given
separation in objective performance.

```{r example-turnover}
turnover <- linkage_turnover(
  solutions,
  objectives = c("cost", "benefit"),
  pairs = neighbors,
  decision_metric = "jaccard"
)

turnover |>
  dplyr::select(
    from_solution,
    to_solution,
    objective_distance,
    decision_distance,
    reconfiguration_rate
  )
```

Informative pairs can then be selected reproducibly with
[`linkage_contrasts()`](https://josesalgr.github.io/multiscape/reference/linkage_contrasts.html).
For example, the contrasts below identify the local transitions with the
largest spatial reconfiguration relative to objective-space change.

```{r example-contrasts}
linkage_contrasts(
  turnover,
  type = "high_reconfiguration",
  n = 2
) |>
  dplyr::select(
    contrast_rank,
    contrast_type,
    from_solution,
    to_solution,
    objective_distance,
    decision_distance,
    reconfiguration_rate
  )
```

`selection_consistency()` provides a complementary decision-space summary of
which planning-unit states recur or vary across the alternatives considered.

